---
title: "Phenotypic Heatmaps & Differential Abundance "
author: "Florian Bach"
date: "08/11/2022"
fig_width: 11
fig_height: 11
---

```{r setup, include=FALSE, results='hide',fig.keep='all'}
knitr::opts_chunk$set(echo = TRUE)
```

## R Phenotypic Heatmaps

The CATALYST package comes with a function called `plotClusterHeatmap`. It's pretty good, but not the most beautiful and you can't easily customise it with info I wanted to add. So what I did is basically look at the source code of `plotClusterHeatmap`, grab what you need to make a heatmap and then you can manually add more layers as you please. The next couple chunks of code will look pretty dense but you don't really need to worry about that, there's just a couple of arguments you need to adapt to your needs based on what you've called your variables before. At first, we will define the functions we need, then we will run them.

The first function I creatively (re)named `shplit_cells` basically splits your dataset into each cluster according to a user-defined clustering, so here you need to put "Meta40" or "Wiebke_merge3" or whatever it is you want.

The second function was named after the sound you make when your code won't run for the 100th time, `ahgg`. This function take the output of `shplit_cells` and calculates summary statistics like median expression (the default) but it can also do mean and sum. It takes 3 arguments: the SCE you're working on, the attributes by which you want to aggregate cells to calculate medians, lastly the summary statistic, usually the median.

```{r defining funs, echo=TRUE, message=FALSE,results='hide', warning=FALSE}

library(CATALYST)
library(diffcyt)
library(SummarizedExperiment)

library(purrr)
library(data.table)
library(reshape2)
library(dplyr)
library(tidyr)
library(stringr)

library(ComplexHeatmap)
library(colorspace)
library(ggplot2)

shplit_cells <- function(x, by) {
  stopifnot(is.character(by), by %in% colnames(colData(x)))
  cd <- data.frame(colData(x))
  cd$cluster_id <- cluster_ids(x, k="flo_merge")
  dt <- data.table::data.table(cd, i = seq_len(ncol(x)))
  dt_split <- split(dt, by = by, sorted = TRUE, flatten = FALSE)
  purrr::map_depth(dt_split, length(by), "i")
}


ahgg <- function(x, by, fun = c("median", "mean", "sum")) {
  fun <- switch(match.arg(fun),
                median = rowMedians, mean = rowMeans, sum = rowSums)
  cs <- shplit_cells(x, by)
  pb <- purrr::map_depth(cs, -1, function(i) {
    if (length(i) == 0) return(numeric(nrow(x)))
    fun(assay(x, "exprs")[, i, drop = FALSE])
  })
  purrr::map_depth(pb, -2, function(u) as.matrix(data.frame(
    u, row.names = rownames(x), check.names = FALSE)))
}

```

Now that we've defined these functions and defined the clustering we want to use, we can calculate the matrix that will become the basis of our phenotypic heatmap. We will first subset our SCE to only include channels of interest that will be columns of our heatmap. What I do is I save the channels in a txt and then load that into my scripts, that way, if you ever change your mind about what channels you want to include, you can simply update that one file and all your scripts will automatically have the correct list, like magic.

I would save the output of this chunk as a csv because it'll be much quicker to just load that file and change your heatmap code downstream, rather than caluclate everything again.

```{r load shit, echo=FALSE}

setwd("~/PhD/cytof/vac69a/reprocessed/reprocessed_relabeled_comped/T_cells_only/")
sce <- vac69a.cytof::read_full("~/PhD/cytof/vac69a/reprocessed/reprocessed_relabeled_comped/T_cells_only/")

```


```{r run funs, echo=TRUE}

refined_markers <- read.csv("~/PhD/cytof/vac69a/reprocessed/reprocessed_relabeled_comped/T_cells_only/refined_markers.csv")$refined_markers

ms <- ahgg(sce[refined_markers,], by = c("cluster_id", "timepoint"), fun="median")

ms2 <- lapply(ms, function(x)data.frame(x))
ms2 <- Map(cbind, ms2, cluster_id = names(ms))

ms3 <- data.table::rbindlist(ms2)
ms3[,Marker := unlist(lapply(ms2, rownames))]

# here we pick the phenotype at T6 to display activated cells in all their glory
ms3 <- select(ms3, T6, cluster_id, Marker)

ms4 <- as.matrix(tidyr::spread(ms3, cluster_id, T6))

rownames(ms4) <- ms4[,1]
ms5 <- ms4[,2:ncol(ms4)]

scaled_mat <- apply(apply(ms5, c(1,2), as.numeric), MARGIN = 1, function(x)scales::rescale(x, to=c(0, 1)))

reordered_scaled_mat <- scaled_mat[,match(refined_markers, colnames(scaled_mat))]


write.csv(scaled_mat, "/home/flobuntu/PhD/cytof/vac63c/normalised_renamed_comped/T_cells_only/cluster_medians_prim_ter_T6.csv")




```

And here's how we make a basic but beautiful phenotypic heatmap. You can subset the matrix to only include specific channels or clusters, just like you would subset a dataframe.

Adding annotations to the heatmap is not the most straightforward thing in the world- you kind of have to hard-code everything because it's all in base R. In a way that's better than ggplot because it's way more flexible and you can make some very bespoke heatmaps, but it's a bit rough if you're not familiar with it. There's lots of resources in the extensive documentation for ComplexHeatmap, so you can have a look there. If there's specific things you need I can help out, but writing up something better than the package mantainer has already provided would be quite a lot of work. That being said, in the differential heatmap section I've included code that I wrote in order to make the annotations, if you want to make exactly those (or very similar ones), it shouldn't be too hard to adapt.



```{r pressure, echo=TRUE}


# bussin color palette; ; you can play with the number of colors you choose here to change the palette slightly
inferno <- colorspace::sequential_hcl("inferno", n=8)

# ComplexHeatmap works best with colorpalettes made in the circlize package (written by the same guy); it basically makes a function that takes a numeric value as input and gives a colour hexcode as output
col_inferno <- circlize::colorRamp2(seq(0,1, by=1/{length(inferno)-1}), inferno)



all_cluster_heatmap <- Heatmap(matrix = as.matrix(reordered_scaled_mat),
                               #cluster_rows = FALSE,
                               cluster_rows = TRUE,
                               show_row_dend = TRUE,
                               show_heatmap_legend = FALSE,
                               name = "Median Marker Expression",
                               cluster_columns = FALSE,
                               row_names_side = "left",
                               row_dend_side = "right",
                               col = col_inferno,
                               rect_gp = gpar(col = "white"),
                               column_names_rot = 45
)

draw(all_cluster_heatmap)

# to save a pdf
# pdf("filepath/graphic_name.pdf", width=8, height=8)
# draw(all_cluster_heatmap, padding = unit(c(2, 25, 2, 15), "mm"))
# dev.off()


# to save a png
# png("filepath/graphic_name.png", width=8, height=8, units = "in", res=400)
# draw(all_cluster_heatmap, padding = unit(c(2, 25, 2, 15), "mm"))
# dev.off()

```
## Differential Abundance


For differential abundance we are going to again rely on our trusty CATALYST package. They in term rely on diffcyt, a package developed in the same lab that incorporates a few flexible linear regression models with SingleCellExperiment. The syntax is a bit confusing if you're not used to using these, but for simple study designs they should be pretty straight forward. Stay away from the mixed effects model framework though, I reported a bug about this ages ago and they still haven' fixed it.

First you need to make a dataframe that contains your experimental info, conditions, volunteers, timepoints, that kind of thing. From this we'll make a design matrix. Rows are samples, columns contain sample property levels. If a condition only has 2 levels (e.g. stim and no_stim) there will only be one column in your matrix. For every sample that's stimulated there will be a 1 and every sample that's not stimulated there will be a 0. For each additional level there will be one more column. This is what's called 'dummy coding'. To familiarise yourself with these tables, have a look at the design object. Here it may be useful to have manually set the levels of all relevant columns in your experimental info dataframe so that baseline comes before diagnosis and T6, for example. 

To make a specific comparison (a contrast) we will communicate with diffcyt by giving it a vector of zeroes and ones that basically tell the algorithm which columns to consider. This approach can build linear models that can ask any question you want, provided you have the right samples. In a simple design like below, you basically set everything to 0 except the factor level you're interested in (in this case T6). Because I've ordered the levels of timepoint properly, this will create a comparison between T6 and baseline

Then it's as simple as running diffcyt, giving it a couple more specifics and Bob's your uncle.


```{r run diffcyt}

# decide on what you'd consider significant
FDR_cutoff <- 0.05

# experimental info
ei <- metadata(sce)$experiment_info

# study design; include all columns that you are interested in as fixed effects
design <- createDesignMatrix(ei, c("timepoint", "volunteer"))

# have a look at the column names;
colnames(design)

# we see that T6 is the 4th column, and because we are interested in how frequencies change across all volunteers between baseline and T6 we'll set only that column to 1, everything else is 0
pairwise_contrast_t6 <- createContrast(c(c(0, 0, 0, 1), rep(0,5)))

da_t6 <- diffcyt(sce,
                 design = design,
                 contrast = pairwise_contrast_t6,
                 analysis_type = "DA",
                 method_DA = "diffcyt-DA-edgeR",
                 clustering_to_use = "flo_merge",
                 verbose = T)


#extract results table
da_t6_res <- data.frame(rowData(da_t6$res))

#count number of significant clusters
table(da_t6_res$p_adj < FDR_cutoff)

# quick (ugly) heatmap to look at the results
plotDiffHeatmap(x=sce, y=da_t6_res, fdr = FDR_cutoff, normalize = TRUE)

# make a boxplot of diffcyt results
# choose log2 fold change & FDR cutoff to choose which clusters to display (but for now I want to show all cluster frequencies)
log_FC <- 0
FDR_cutoff <- 1

clusters_that_change <- da_t6_res %>%
  filter(abs(logFC) > log_FC) %>%
  filter(p_adj < FDR_cutoff) %>%
  select(cluster_id)
#pivot_longer(sig_results, cols=colnames(sig_results)[4:ncol(sig_results)], names_to = "sample_id", values_to = "count")


# get table that contains cluster counts, subset so only significant clusters are in
results_table <- data.frame(topTable(da_t6, all=TRUE, show_counts = TRUE))
results_cut <- subset(results_table, results_table$cluster_id %in% clusters_that_change$cluster_id)
  
long_sig_results <- pivot_longer(results_cut, cols=colnames(results_cut)[4:ncol(results_cut)], names_to = "sample_id", values_to = "count")
# for some reason that adds the word counts to the sample id, let's just get rid of that quickly
long_sig_results$sample_id <- gsub("counts_", "", long_sig_results$sample_id)

# in my data i can add volunteer and timepoint columns like this, depending on how you named them, your mileage may vary; i slice the sample name after the occurence of "v" followed by a number
long_sig_results$volunteer <- str_match(long_sig_results$sample_id, "v[0-9]*")[, 1]
# all volunteer names are 3 characters long plus an underscore so character 5 onwards contains the timepoint
long_sig_results$timepoint <- substr(long_sig_results$sample_id, 5,nchar(long_sig_results$sample_id))
  
# get the overall cell counts to calculate percentages
cell_counts <- n_cells(sce)
long_sig_results$frequency <- long_sig_results$count / cell_counts[long_sig_results$sample_id] *100
  
  
time_col <- sequential_hcl(5, palette = "Purple Yellow")
  
all_cluster_boxplot <- ggplot(long_sig_results, aes(x=factor(timepoint), y=frequency))+
    geom_boxplot(aes(fill=timepoint), outlier.shape = NA)+
    geom_point(aes(shape=volunteer))+
    facet_wrap(~cluster_id, scales = "free", ncol=7, labeller=label_wrap_gen(width = 20))+
    theme_minimal()+
    scale_x_discrete(name = "Timepoint")+
    scale_fill_manual(values = c("Baseline"=time_col[4], "C10"=time_col[3], "Diagnosis"=time_col[2], "T6"=time_col[1]))+
    theme(axis.title = element_text(size=10),
          legend.title = element_blank(),
          axis.text.x = element_text(size=8, angle = 45, hjust = 1),
          strip.text = element_text(size=9))
  




# save the file as a csv so we can edit the heatmap without running all this analysis again
# write.csv(long_sig_results, "/home/flobuntu/PhD/cytof/vac69a/reprocessed/reprocessed_relabeled_comped/T_cells_only/all_t6_data.csv")

# also save edgeR results table
# write.csv(da_t6_res, "/home/flobuntu/PhD/cytof/vac69a/reprocessed/reprocessed_relabeled_comped/T_cells_only/da_t6_res.csv")

```
Lastly, to make beautiful differential abundance heatmaps we're taking a similar approach to earlier. So first we need to personally generate the data that goes into it. We're going to be displaying frequency z scores, so we'll have to transform the percentages we generated earlier. Then we convert that into a matrix, choose the clusters to display and could call it quits there for a simple z score heatmap. However, I've included a couple of sections for how to make different kinds of heatmap annotations to display fold change, p value, batch/timepoint information and all that. It's very wordy, but as far as I can see that's the way to make them, a lot of stuff is just hardcoded (and thus very flexible, but also very finicky).

```{r making a heatmap}

# make a transformer
asinTransform <- function(x){asin(sqrt(x))}

# apply transformer cluster-wise and then make z scores
t6_map_data <- long_sig_results%>%
  group_by(cluster_id) %>%
  #group_by(volunteer) %>%
  mutate(trans_freq=asin(sqrt(frequency/100))) %>%
  mutate(max_freq=max(frequency)) %>%
  mutate(trans_norm_freq=scale(trans_freq, center = TRUE, scale = TRUE)) %>%
  ungroup()




# construct matrix to be plotted in heatmap: order by timpoint all baseline samples, dod samples etc. are disaplyed next to each other later on; pivot wider so we have clusters along rows, samples along columns, values=trans_norm_freq
hm_matrix <- as.data.frame(t6_map_data %>%
                             arrange(timepoint)%>%
                             select(cluster_id, sample_id, trans_norm_freq) %>%
                             tidyr::pivot_wider(names_from = sample_id, values_from = trans_norm_freq))

# add rownames (see why in next step)
rownames(hm_matrix) <- hm_matrix$cluster_id

# matrix can only contain numerical data
hm_matrix <- select(hm_matrix, -cluster_id)

# the diffcyt results table should be ordered by fold change, so we can grab its cluster ID column to sort our heatmap by fold change
da_t6_res <- da_t6_res[order(da_t6_res$logFC, decreasing = TRUE),]
cluster_levels <- da_t6_res$cluster_id
hm_matrix <- as.matrix(hm_matrix[match(cluster_levels, rownames(hm_matrix)),])

#truncate matrix values so that extreme values don't skew the colour palette to pale everything else
hm_matrix <- ifelse(hm_matrix> 2.5, 2.5, hm_matrix)
hm_matrix <- ifelse(hm_matrix < -2.5, -2.5, hm_matrix)
extreme <- max(abs(range(hm_matrix)))

# you have to do this, otherwise ComplexHeatmap will likely complain
class(hm_matrix) <- "numeric"

#bunch of some color palettes; the first is written so it interpolates between two colours for high and low fold changes and black around 0
col_fun4 <- circlize::colorRamp2(c(min(hm_matrix), 0, max(hm_matrix)), c("#0859C6", "black", "#FFA500"))
time_col <- colorspace::sequential_hcl(5, palette = "Purple Yellow")

# slice matrix to display the 12 most up/down clusters or however many you want
top_mat <- head(hm_matrix, n=12)
bot_mat <- tail(hm_matrix, n=12)
combo_matrix <- rbind(top_mat, bot_mat)

# let's have a look at what we've made so far

Heatmap(matrix = combo_matrix,
        cluster_rows = FALSE,
        name = "Normalised\nFrequency",
        cluster_columns = FALSE,
        row_names_side = "left",
        column_names_gp = gpar(fontsize=10),
        row_names_gp = gpar(fontsize=10),
        col = col_fun4,
        # this line splits the heatmap into up and down, uncomment if you want to
        # row_split = factor(rep(c("up", "down"), each = 12), levels = c("up", "down")),
        # if you dont want the labels on the side to indicate up and down, uncomment the next line also
        # row_title = c("",""),
        rect_gp = gpar(col = "white"),
        show_heatmap_legend = TRUE,
        column_names_rot = 45,
        heatmap_legend_param = list(col = col_fun4, title = "Z-Score", title_position = "topleft")
)
```

So far so good! The heatmap looks nice and contains everything essential. If you want to add heatmap annotations, here's what I wrote to make the heatmap that we put into the vivax paper.

```{r, heatmap anno, R.options = list(width=2000)}

# make separate vectors for p_adj, significance, fold change etc for heatmap annotations

# right annotation: display fold change between baseline and T6; grab highest and lowest values, exactly as often as we need
p_adj <- c(head(da_t6_res$p_adj, n=nrow(top_mat)), tail(da_t6_res$p_adj, n=nrow(bot_mat)))
log2_fc <- c(head(da_t6_res$logFC, n=nrow(top_mat)), tail(da_t6_res$logFC, n=nrow(bot_mat)))

# left annotation: stats results
significant <- ifelse((p_adj<0.05 & abs(log2_fc)>1), "<0.05", ">0.05")
sig <- c("<0.05"="darkgreen", ">0.05"="lightgrey")

#top annotation; hardcode volunteer colours as list
Volunteer <- c("v02" = "#FB9A99","v03" = "#E31A1C","v05" = "#A6CEE3", "v06" = "#1F78B4", "v07" = "#F0E442", "v09" = "#E69F00")
#hardcode timepoint colours as list
Timepoint <- c("Baseline"=time_col[4], "C10"=time_col[3], "Diagnosis"=time_col[2], "T6"=time_col[1])
#hardcode batch colours as list
Batch <- c("batch_1"="lightgrey", "batch_2"="darkgrey")


# make barplot annotation and include stylistic details
combo_right_anno <-  rowAnnotation(gap = unit(2, "mm"),
                                   "log2FC" = anno_barplot(log2_fc, which="row", axis = TRUE, ylim = c(-2, 6), axis_param = list(at=seq(-2, 6, by=2)),gp = gpar(fill = col_fun4(log2_fc))),
                                   width = unit(3, "cm") # width of the line graph
)

# make left annotation and add stylistic details
combo_left_anno_var <-  rowAnnotation(gap = unit(5, "mm"),
                                      #annotation_name_gp = gpar(angle=45),
                                      show_annotation_name = FALSE,
                                      "FDR"=significant,
                                      simple_anno_size = unit(2.5, "mm"), # width of the significance bar
                                      col=list("FDR" = c("<0.05"="darkgreen", ">0.05"="lightgrey")),
                                      annotation_legend_param = list(FDR = list(title ="FDR",
                                                                                        at = rev(names(sig)),
                                                                                        #title_gp=gpar(angle=45),
                                                                                        legend_gp = gpar(fill = unname(sig)),
                                                                                        title_position = "topleft")
                                      )
                                      
)


#annotation_name_gp = gpar(fontsize=10)
# make top annotation which contains a bunch of info related to the samples, so there's a lot to be done manually
combo_top_anno <- HeatmapAnnotation(gap = unit(2, "mm"), annotation_name_side = "left", annotation_name_gp = gpar(fontsize=10),
                                    Volunteer = rep(levels(ei$volunteer), 4),
                                    Timepoint = rep(levels(ei$timepoint), each=6),
                                    Batch = rep(rep(levels(ei$batch), each=3), 4),
                                    col=list(Batch = c("batch_1"="lightgrey", "batch_2"="darkgrey"),
                                             Timepoint = c("Baseline"=time_col[4], "C10"=time_col[3], "Diagnosis"=time_col[2], "T6"=time_col[1]),
                                             Volunteer = c("v02" = "#FB9A99","v03" = "#E31A1C","v05" = "#A6CEE3", "v06" = "#1F78B4", "v07" = "#F0E442", "v09" = "#E69F00")
                                    ),
                                    simple_anno_size = unit(2.5, "mm"),
                                    annotation_legend_param = list(
                                      Volunteer = list(title = "Volunteer", at = names(Volunteer), legend_gp = gpar(fill = unname(Volunteer)), title_position = "topleft"),
                                      Timepoint = list(title ="Timepoint",at = names(Timepoint), legend_gp = gpar(fill = unname(Timepoint)), title_position = "topleft"),
                                      Batch = list(title = "Batch", at = names(Batch), legend_gp = gpar(fill = unname(Batch)), title_position = "topleft")
                                    )
)


# get rid of underscores in column names, change gamma delta cluster names
colnames(combo_matrix) <- gsub("_", " ", colnames(combo_matrix))
rownames(combo_matrix) <- gsub("gamma delta","gd", rownames(combo_matrix), fixed = T)  

Heatmap(matrix = combo_matrix,
                     cluster_rows = FALSE,
                     name = "Normalised\nFrequency",
                     cluster_columns = FALSE,
                     row_names_side = "left",
                     column_names_gp = gpar(fontsize=10),
                     row_names_gp = gpar(fontsize=10),
                     col = col_fun4,
                     row_split = factor(rep(c("up", "down"), each = 12), levels = c("up", "down")),
                     rect_gp = gpar(col = "white"),
                     row_title = c("",""),
                     top_annotation = combo_top_anno,
                     right_annotation = combo_right_anno,
                     left_annotation = combo_left_anno_var,
                     show_heatmap_legend = TRUE,
                     column_names_rot = 45,
                     heatmap_legend_param = list(col = col_fun4, title = "Z-Score", title_position = "topleft")
)



```

<!-- Et voilà! Fancy heatmap achieved. -->
