---
title: "Goncalves Modelling (Model Selection via Maximum Likelihood)"
author: "Florian Bach"
date: "21/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="~/postdoc/goncalves_modelling/markdown")


library(dplyr, quietly=TRUE)
library(ggplot2, quietly=TRUE)
library(stats4, quietly = TRUE)
```


## Hypotheses

To further characterise the the best model to describe the risk dynamics of severe disease we will use maximum likelihood estimation and AIC to test several models:

1. The risk of severe disease is constant across infections
2. The risk of severe disease declines at a constant rate with the number of infections.
3. The risk of severe disease  drops sharply from 1st to 2nd infection in life and then declines at a constant rate.
4. The risk of severe disease declines exponentially.


As a reminder, on the left are the fractions we're modelling, which now include both moderate as well as severe cases. To deal with missing case numbers where no cases of moderate/severe disease were recorded, I used weighted linear regression of non-`NA`, log-transformed case counts to model the number of cases at each infection (dashed line) and imputed the rounded values. 

```{r severe probability plot, echo=FALSE, warning=FALSE, fig.show="hold", out.width="50%"}

goncalves_data <- data.frame("N_Infection"=seq(0,14),
                             "Severe"=c(0, 32, 18, 20, 9, 9, 2, 2, 8, NA, 1, NA, NA, NA, 1),
                             "Infections"=c(0, 715, 540, 424, 330, 267, 220, 168, 131, NA, 90, NA, NA, NA, 25),
                             "Susceptible"=c(102, 70, 52, 32, 23, 14, 12, 10, 2, 2, 1, 1, 1, 1, 0),
                             "Severe_Moderate"=c(0, 76, 41, 31, 15, 12, 5, 8, 5, 1, 1, 3, NA, NA, 1),
                             "Supp_Infections"=c(0, 715, 501, 369, 276, 213, 169, 120, 86, 62, 55, 47, NA, NA, 25))

goncalves_data$Severe_Moderate <- ifelse(is.na(goncalves_data$Severe_Moderate), 0, goncalves_data$Severe_Moderate)

goncalves_data$Severe <- ifelse(is.na(goncalves_data$Severe), 0, goncalves_data$Severe)

# goncalves_plot_data <- subset(goncalves_data, goncalves_data$N_Infection>0&goncalves_data$N_Infection<12)

supp_infection_decay <- lm(log(Supp_Infections)~N_Infection, data=goncalves_data[2:15,], weights = Supp_Infections)

supp_prd <- data.frame(N_Infection = seq(from = 1, to = 14), length.out=1000)
supp_err <- predict(supp_infection_decay, newdata = supp_prd, se.fit = TRUE)
supp_prd$modelled_cases <- exp(supp_err$fit)
supp_prd$round_model <- round(supp_prd$modelled_cases)

goncalves_data$Modelled_Supp_Cases <- ifelse(is.na(goncalves_data$Supp_Infections), supp_prd$round_model, goncalves_data$Supp_Infections)

goncalves_data$Supp_Imputed <- ifelse(is.na(goncalves_data$Supp_Infections), TRUE, FALSE)




main_infection_decay <- lm(log(Infections)~N_Infection, data=goncalves_data[2:15,], weights = Infections)

main_prd <- data.frame(N_Infection = seq(from = 1, to = 14), length.out=100)
main_err <- predict(main_infection_decay, newdata = main_prd, se.fit = TRUE)
main_prd$modelled_cases <- exp(main_err$fit)
main_prd$round_model <- round(main_prd$modelled_cases)

goncalves_data$Modelled_Main_Cases <- ifelse(is.na(goncalves_data$Infections), main_prd$round_model, goncalves_data$Infections)

goncalves_data$Main_Imputed <- ifelse(is.na(goncalves_data$Infections), TRUE, FALSE)





goncalves_plot_data <- goncalves_data


moderate_disease_bar <- ggplot(goncalves_data, aes(x = N_Infection, y=Severe_Moderate/Modelled_Supp_Cases))+
  ylab("\n\nRisk of Severe / Moderate Disease\n")+
  scale_y_continuous(label=scales::percent, limits=c(0, 0.13))+
  scale_x_continuous(breaks=seq(1, nrow(goncalves_plot_data)), expand = expansion(add = c(0.2,0)))+
  geom_text(aes(label= paste0("frac(",Severe_Moderate, ",", Modelled_Supp_Cases,")")),parse = TRUE, vjust= -0.2, data = goncalves_plot_data, size=3.5)+
  geom_bar(stat="identity", fill="darkblue", na.rm = TRUE)+
  ggtitle("Severe / Moderate Disease")+
  scale_fill_manual(values = c("darkblue", "blue"))+
  theme_minimal()+
  theme(plot.title = element_text(size=15, hjust = 0.5),
        axis.text = element_text(size=12),
        axis.title = element_text(size=13.5)
        #plot.margin = unit(c(2,2,2,2), "lines")
        )

supp_case_decay <- ggplot()+
  geom_point(data=goncalves_plot_data, colour="darkblue", aes(x=N_Infection, y=Modelled_Supp_Cases, shape=Supp_Imputed))+
geom_line(data = supp_prd, aes(x = N_Infection, y=round_model), colour="blue", linetype="dashed")+
  ylab("\n\nNumber of Cases\n")+
  scale_x_continuous(breaks=seq(1, nrow(goncalves_plot_data)-1),
                     limits=range(1,nrow(goncalves_plot_data)-1),
                     expand = expansion(add = c(0.2,0.2)))+
  ggtitle("Number of Cases with Imputed Values")+
  theme_minimal()+
  guides(shape=guide_legend(title="Imputed"))+
  theme(plot.title = element_text(size=15, hjust = 0.5),
        axis.text = element_text(size=12),
        axis.title = element_text(size=13.5)
        #plot.margin = unit(c(2,2,2,2), "lines")
        )




severe_disease_bar <- ggplot(goncalves_data, aes(x = N_Infection, y=Severe/Modelled_Main_Cases))+
  ylab("\n\nRisk of Severe Disease\n")+
  scale_y_continuous(label=scales::percent, limits=c(0, 0.13))+
  scale_x_continuous(breaks=seq(1, nrow(goncalves_plot_data)), expand = expansion(add = c(0.2,0)))+
  geom_text(aes(label= paste0("frac(",Severe, ",", Modelled_Main_Cases,")")),parse = TRUE, vjust= -0.2, data = goncalves_plot_data, size=3.5)+
  geom_bar(stat="identity", fill="darkred", na.rm = TRUE)+
  ggtitle("Severe Disease")+
  scale_fill_manual(values = c("darkred", "red"))+
  theme_minimal()+
  theme(plot.title = element_text(size=15, hjust = 0.5),
        axis.text = element_text(size=12),
        axis.title = element_text(size=13.5)
        #plot.margin = unit(c(2,2,2,2), "lines")
        )

main_case_decay <- ggplot()+
  geom_point(data=goncalves_plot_data, colour="darkred", aes(x=N_Infection, y=Modelled_Main_Cases, shape=Main_Imputed))+
geom_line(data = main_prd, aes(x = N_Infection, y=round_model), colour="red", linetype="dashed")+
  ylab("\n\nNumber of Cases\n")+
  scale_x_continuous(breaks=seq(1, nrow(goncalves_plot_data)-1),
                     limits=range(1,nrow(goncalves_plot_data)-1),
                     expand = expansion(add = c(0.2,0.2)))+
  ggtitle("Number of Cases with Imputed Values")+
  theme_minimal()+
  guides(shape=guide_legend(title="Imputed"))+
  theme(plot.title = element_text(size=15, hjust = 0.5),
        axis.text = element_text(size=12),
        axis.title = element_text(size=13.5)
        #plot.margin = unit(c(2,2,2,2), "lines")
        )






#combo_plot <- cowplot::plot_grid(severe_disease_bar, moderate_disease_bar)

plot(moderate_disease_bar)
plot(supp_case_decay)

plot(severe_disease_bar)
plot(main_case_decay)
```


## Model Definitions
  
We will model the data as derived from a binomial distribution $N \sim {\sf Binom}(M, P_{i})$ where

* $N$ is a vector of successes (i.e. instances of severe disease at each order of infection $i$)
* $M$ is a vector of the number of observations (i.e. number infections at each order of infection $i$)
* $P_{i}$ is a vector of probabilities giving the probability $p$ of success at each trial (i.e. the severe disease risk at each order of infection $i$)

The first two parameters $N, M$ are fixed, as they correspond to the numerator and denominator of the fractions of severe disease, respectively, at each order of infection. The vector of probabilities of success, $P_{i}$, is what we aim to estimate using maximum likelihood. In this format the above hypotheses 1-4) can be rewritten as estimating $P_{i}$ as:

1. $P_{i} = p_{1}$ (constant for all $i$)
2. $P_{i} = p_{1}-\alpha(i-1)$
3.  *   $P_{i} = p_{1} - \alpha(i-1)$ for $i \leqq j$
    *   $P_{i} = p_{1} - \alpha(j-1) - \alpha_2 (i-j)$ for $i > j$
4. $P_{i} = p_{1} e^{-\alpha(i-1)}$



For determining the maximum likelihood I used `stats4::mle()` which implements the Nelder-Mead algorithm (https://doi.org/10.1093/comjnl/7.4.308).

```{r defining binom functions for moderate, echo=FALSE, warning=FALSE}

# moderate severity

n_successes <- goncalves_data$Severe_Moderate[2:15]
m_trials <- goncalves_data$Modelled_Supp_Cases[2:15]

# n_successes <- goncalves_data$Severe[2:9]
# m_trials <- goncalves_data$Infections[2:9]


## flat probability

# this returns a log likelihood; setting log = FALSE and multiplying returns the same , non-log result
flat_likelihood <- function(p){
  sum(dbinom(n_successes, m_trials, p, log = TRUE))*-1
}

no_decay_estimate <- stats4::mle(minuslog=flat_likelihood, start=list(p=0.1))

# logLik(no_decay_estimate)
# summary(no_decay_estimate)
# log Lik.' -27.72533 (df=1)
# Coefficients:
#     Estimate  Std. Error
# p 0.07880817 0.005443811
# 
# -2 log L: 55.45065




## steady decay

# alpha cannot exceed the fraction of p over the number of observations (otherwise we may venture into negative
# probability space) 

steady_decay_likelihood <- function(p_i, alpha){
  p_vec <- p_i - alpha * (seq(1, length(n_successes), by=1)-1)
  res <- sum(
    dbinom(x = n_successes, size = m_trials, prob = p_vec, log = TRUE)
    )*-1
  res <- ifelse(is.na(res), 100, res)
  return(res)
}

steady_decay_estimate <- stats4::mle(minuslog=steady_decay_likelihood, start=list(p_i=0.05, alpha=0.001))

# logLik(steady_decay_estimate)
# summary(steady_decay_estimate)
# Coefficients:
#          Estimate  Std. Error
# p_i   0.096175803 0.007841390
# alpha 0.008368463 0.002122148
# 
#'log Lik.' -21.54903 (df=2)


## exponential decay

exponential_decay_likelihood <- function(p_i, alpha){
  p_vec <- p_i * exp(- alpha * (seq(1, length(n_successes), by=1)-1))
  
  res <- sum(
    dbinom(x = n_successes, size = m_trials, prob = p_vec, log = TRUE)
    )*-1
  
  res <- ifelse(is.na(res), 100, res)
}

exponential_decay_estimate <- stats4::mle(minuslog=exponential_decay_likelihood, start=list(p_i=0.1,alpha=0.01))

# logLik(exponential_decay_estimate)
# summary(exponential_decay_estimate)
# Coefficients:
#        Estimate  Std. Error
# p_i   0.1013627 0.009283059
# alpha 0.1386613 0.039169689
# 
#'log Lik.' -20.78666 (df=2)

## stepwise decay

# nick_decay_likelihood2 <- function(p_i, alpha){
#   
#   p_vec <- p_i - alpha * (seq(1, length(n_successes), by=1)-3)
#   sum(
#     dbinom(x = n_successes, size = m_trials, prob = p_vec, log = TRUE)
#     )*-1
# }


nick_decay_likelihood <- function(p_i, alpha_1, alpha_2, i){

     #define stepwise function
  p_vec_1 <- p_i - alpha_1 * 0:length(n_successes)
  p_vec_2 <- p_i - alpha_1 * 0:length(n_successes) - alpha_2 * (1:length(n_successes)+1-round(i))
     
  #combine based on nick position
  p_vec <- c(p_vec_1[1:round(i)], p_vec_2[(round(i)+1):length(n_successes)])
  
  fail_safe <- ifelse(any(p_vec<0), TRUE, FALSE)
  p_vec <- ifelse(p_vec<0, 0.00000001, p_vec)
  # 
  #calc distribution
  res <- ifelse(i%%1==0,
               sum(
                   dbinom(x = n_successes, size = m_trials, prob = p_vec, log = TRUE)
                           )*-1,
                               100)
       
        
  res <- ifelse(fail_safe==TRUE, 100, res)
  res <- ifelse(is.na(res), 100, res)
  res <- ifelse(is.infinite(res), 100, res)

  return(res)}
   
   
moderate_brute_force <- data.frame(matrix(ncol=9))

colnames(moderate_brute_force) <- c("P1", "SE(P1)", "alpha_1", "SE(alpha_1)", 	"alpha_2", 	"SE(alpha_2)", "j", "logL", "AIC")
   
   
for(number in 2:8){
  
  nick_decay_estimate <- stats4::mle(#upper = c(0.15, 0.05, 0.01, number),
                                     lower = c(0.05, 0, -0.02, number),
                                     minuslog = nick_decay_likelihood,
                                     start = list(p_i=0.1,
                                                  alpha_1=0,
                                                  alpha_2=0,
                                                  i=number)
                                                  )
           
  #print(nick_decay_estimate@vcov)
  
     
  new_coefs <- summary(nick_decay_estimate)@coef[,1]
  new_se <- summary(nick_decay_estimate)@coef[,2]
             
  new_entry <- vector(length = 9)
               
  new_entry[c(1,3,5,7)] <- summary(nick_decay_estimate)@coef[,1]
  new_entry[c(2,4,6)] <- summary(nick_decay_estimate)@coef[1:3,2]
  new_entry[8] <- logLik(nick_decay_estimate)
  new_entry[9] <- AIC(nick_decay_estimate)
               
                   
  moderate_brute_force[number-1,] <- new_entry
}
 



# glm 


moderate_fit <- glm(Severe_Moderate/Modelled_Supp_Cases~N_Infection, weights = Modelled_Supp_Cases, family="binomial", data=goncalves_plot_data)

intercept_se_calculator <- function(x){
  uci=exp(summary(x)$coef[1]+(1.96*summary(x)$coef[3]))
  lci=exp(summary(x)$coef[1]-(1.96*summary(x)$coef[3]))
  se=(uci-lci)/3.92
  return(se)
}

slope_se_calculator <- function(x){
  uci=exp(summary(x)$coef[2]+(1.96*summary(x)$coef[4]))
  lci=exp(summary(x)$coef[2]-(1.96*summary(x)$coef[4]))
  se=(uci-lci)/3.92
  return(se)
}



moderate_results_df <- data.frame(check.names = FALSE,
                         "Model"=c("No Decay",
                                "Constant Decay",
                                #"Constant Decay with Kink",
                                "Exponential Decay",
                                "Exponential Decay (glm)"),
                      "P1"=c(no_decay_estimate@coef,
                              steady_decay_estimate@coef[1],
                              #nick_2decay_estimate@coef[1],
                              exponential_decay_estimate@coef[1],
                             exp(moderate_fit$coefficients[1])),
                      "SE(P1)"=c(summary(no_decay_estimate)@coef[1,2],
                                  summary(steady_decay_estimate)@coef[1,2],
                                  #summary(nick_2decay_estimate)@coef[1,2],
                                  summary(exponential_decay_estimate)@coef[1,2],
                                  intercept_se_calculator(moderate_fit)),
                      "alpha"=c(NA,
                              steady_decay_estimate@coef[2],
                              #nick_2decay_estimate@coef[2],
                              exponential_decay_estimate@coef[2],
                              (-moderate_fit$coefficients[2])),
                      "SE(alpha)"=c(NA,
                                  summary(steady_decay_estimate)@coef[2,2],
                                  #summary(nick_2decay_estimate)@coef[2,2],
                                  summary(exponential_decay_estimate)@coef[2,2],
                                  summary(moderate_fit)$coef[2,2]),
                      "logL"=c(logLik(no_decay_estimate),
                               logLik(steady_decay_estimate),
                               #logLik(nick_2decay_estimate),
                               logLik(exponential_decay_estimate),
                               logLik(moderate_fit)),
                      "AIC"=c(AIC(no_decay_estimate),
                               AIC(steady_decay_estimate),
                               #AIC(nick_2decay_estimate),
                               AIC(exponential_decay_estimate),
                               AIC(moderate_fit))
                               
                      )
```



```{r defining binom functions for severe, echo=FALSE, warning=FALSE}

n_successes <- goncalves_data$Severe[2:15]
m_trials <- goncalves_data$Modelled_Main_Cases[2:15]

## flat probability

# this returns a log likelihood; setting log = FALSE and multiplying returns the same , non-log result
flat_likelihood <- function(p){
  sum(dbinom(n_successes, m_trials, p, log = TRUE))*-1
}

no_decay_estimate <- stats4::mle(minuslog=flat_likelihood, start=list(p=0.1))



## steady decay

# alpha cannot exceed the fraction of p over the number of observations (otherwise we may venture into negative
# probability space) 

steady_decay_likelihood <- function(p_i, alpha){
  p_vec <- p_i - alpha * (seq(1, length(n_successes), by=1)-1)
  res <- sum(
    dbinom(x = n_successes, size = m_trials, prob = p_vec, log = TRUE)
    )*-1
  res <- ifelse(is.na(res), 100, res)
  return(res)
}

steady_decay_estimate <- stats4::mle(minuslog=steady_decay_likelihood, start=list(p_i=0.05, alpha=0))

# logLik(steady_decay_estimate)
# summary(steady_decay_estimate)
# Coefficients:
#          Estimate  Std. Error
# p_i   0.096175803 0.007841390
# alpha 0.008368463 0.002122148
# 
#'log Lik.' -21.54903 (df=2)


## exponential decay

exponential_decay_likelihood <- function(p_i, alpha){
  p_vec <- p_i * exp(- alpha * (seq(1, length(n_successes), by=1)-1))
  
  res <- sum(
    dbinom(x = n_successes, size = m_trials, prob = p_vec, log = TRUE)
    )*-1
  
  res <- ifelse(is.na(res), 100, res)
}

exponential_decay_estimate <- stats4::mle(minuslog=exponential_decay_likelihood, start=list(p_i=0.01,alpha=0.01))

# logLik(exponential_decay_estimate)
# summary(exponential_decay_estimate)
# Coefficients:
#        Estimate  Std. Error
# p_i   0.1013627 0.009283059
# alpha 0.1386613 0.039169689
# 
#'log Lik.' -20.78666 (df=2)

## stepwise decay

# nick_decay_likelihood2 <- function(p_i, alpha){
#   
#   p_vec <- p_i - alpha * (seq(1, length(n_successes), by=1)-3)
#   sum(
#     dbinom(x = n_successes, size = m_trials, prob = p_vec, log = TRUE)
#     )*-1
# }


nick_decay_likelihood <- function(p_i, alpha_1, alpha_2, i){

     #define stepwise function
  p_vec_1 <- p_i - alpha_1 * 0:length(n_successes)
  p_vec_2 <- p_i - alpha_1 * 0:length(n_successes) - alpha_2 * (1:length(n_successes)+1-round(i))
     
  #combine based on nick position
  p_vec <- c(p_vec_1[1:round(i)], p_vec_2[(round(i)+1):length(n_successes)])
  
  fail_safe <- ifelse(any(p_vec<0), TRUE, FALSE)
  p_vec <- ifelse(p_vec<0, 0.00000001, p_vec)
  # 
  #calc distribution
  res <- ifelse(i%%1==0,
               sum(
                   dbinom(x = n_successes, size = m_trials, prob = p_vec, log = TRUE)
                           )*-1,
                               200)
       
        
  res <- ifelse(fail_safe==TRUE, 200, res)
  res <- ifelse(is.na(res), 200, res)
  res <- ifelse(is.infinite(res), 200, res)

  return(res)}
   
 
# severe_brute_force <- data.frame(matrix(ncol=9))
# 
# colnames(severe_brute_force) <- c("P1", "SE(P1)", "alpha_1", "SE(alpha_1)", 	"alpha_2", 	"SE(alpha_2)", "j", "logL", "AIC")
#    
#    
# for(number in 2:8){
#   
#   nick_decay_estimate <- stats4::mle(#upper = c(0.15, 0.05, 0.01, number),
#                                      #lower = c(0.0001, 0.0, -0.20, number),
#                                      minuslog = nick_decay_likelihood, fixed = list(i=number),
#                                      start = list(p_i=0,
#                                                   alpha_1=0,
#                                                   alpha_2=0,
#                                                   i=number)
#                                                   )
#            
#   #print(nick_decay_estimate@vcov)
#   
#      
#   new_coefs <- summary(nick_decay_estimate)@coef[,1]
#   new_se <- summary(nick_decay_estimate)@coef[,2]
#              
#   new_entry <- vector(length = 9)
#                
#   new_entry[c(1,3,5,7)] <- summary(nick_decay_estimate)@coef[,1]
#   new_entry[c(2,4,6)] <- summary(nick_decay_estimate)@coef[1:3,2]
#   new_entry[8] <- logLik(nick_decay_estimate)
#   new_entry[9] <- AIC(nick_decay_estimate)
#                
#                    
#   severe_brute_force[number-1,] <- new_entry
# }
#  
# 

severe_fit <- glm(Severe/Modelled_Main_Cases~N_Infection, weights = Modelled_Main_Cases, family="binomial", data=goncalves_plot_data)



severe_results_df <- data.frame(check.names = FALSE,
                         "Model"=c("No Decay",
                                "Constant Decay",
                                #"Constant Decay with Kink",
                                "Exponential Decay",
                                "Exponential Decay (glm)"),
                      "P1"=c(no_decay_estimate@coef,
                              steady_decay_estimate@coef[1],
                              #nick_2decay_estimate@coef[1],
                              exponential_decay_estimate@coef[1],
                             exp(severe_fit$coefficients[1])),
                      "SE(P1)"=c(summary(no_decay_estimate)@coef[1,2],
                                  summary(steady_decay_estimate)@coef[1,2],
                                  #summary(nick_2decay_estimate)@coef[1,2],
                                  summary(exponential_decay_estimate)@coef[1,2],
                                  intercept_se_calculator(severe_fit)),
                      "alpha"=c(NA,
                              steady_decay_estimate@coef[2],
                              #nick_2decay_estimate@coef[2],
                              exponential_decay_estimate@coef[2],
                              (-severe_fit$coefficients[2])),
                      "SE(alpha)"=c(summary(no_decay_estimate)@coef[,2],
                                  summary(steady_decay_estimate)@coef[2,2],
                                  #summary(nick_2decay_estimate)@coef[2,2],
                                  summary(exponential_decay_estimate)@coef[2,2],
                                  summary(severe_fit)$coef[2,2]),
                      "logL"=c(logLik(no_decay_estimate),
                               logLik(steady_decay_estimate),
                               #logLik(nick_2decay_estimate),
                               logLik(exponential_decay_estimate),
                               logLik(severe_fit)),
                      "AIC"=c(AIC(no_decay_estimate),
                               AIC(steady_decay_estimate),
                               #AIC(nick_2decay_estimate),
                               AIC(exponential_decay_estimate),
                               AIC(severe_fit))
                               
                      )

```

```{r results table, echo=FALSE}

knitr::kable(moderate_results_df, caption="Model Comparison for Moderate Disease") %>%
  kableExtra::kable_styling(full_width = TRUE)

knitr::kable(severe_results_df, caption="Model Comparison for Severe Disease") %>%
  kableExtra::kable_styling(full_width = TRUE)


# 
# knitr::kable(moderate_brute_force, caption="Stepwise Models for Moderate Disease") %>%
#   kableExtra::kable_styling(full_width = TRUE)


```
The stepwise models were problematic for modelling severe disease and given their poor performance for moderate disease I decided to not devote any more resources to this at this stage and just leave them out of this report for now. 

Based on AIC & log likelihood, exponential decay provides the best fit for moderate disease data. The logistic regression model suggested by `stats::glm()` is shown below.  

For severe cases, constant and exponential decay had similar likelihood and AIC and both models are of equal complexity. It is therefore not immediately obvious which should be chosen and both are shown below. I would make the case that since moderate cases were best described by exponential decay, so should severe cases. In any case, imputing the missing case numbers improved model fit such that a "no decay" model no longer provides an equal or better fit, suggesting that the risk of severe malaria does indeed decrease over time, which is the more important observation, be it linear or exponential. And it is contrary to what the authors claimed in the original publication. It would be interesting to see how this holds up once we have the real number of observed cases, but I'm optimistic.

For getting in touch with Bronner and/or Patrick I would say that the minimum information required that would be useful for advancing this analysis is therefore the missing denominators, as well as some information on what number of children with "moderate disease" were hospitalised at each order of infection. Individual data would of course be ideal for more complex models that can correct for relevant confounders, but this would of course involve a much larger amount of work and commitment.


```{r function plots, warning=FALSE, echo=FALSE, out.width="66.6%", fig.align='center'}
severe_glm <- glm(Severe/Modelled_Main_Cases~N_Infection, weights = Modelled_Main_Cases, family="binomial", data=goncalves_plot_data)
severe_exp_fun <- function(x){exp(severe_glm$coefficients[1])*exp(severe_glm$coefficients[2])^x}

severe_constant_fun <- function(x){steady_decay_estimate@coef[1]-steady_decay_estimate@coef[2]*x}

moderate_glm <- glm(Severe_Moderate/Modelled_Supp_Cases~N_Infection, weights = Modelled_Supp_Cases, family="binomial", data=goncalves_plot_data)
moderate_exp_fun <- function(x){exp(moderate_glm$coefficients[1])*exp(moderate_glm$coefficients[2])^x}


prd <- data.frame(N_Infection = seq(from = 1, to = 14, length.out = 100))
err <- predict(moderate_glm, newdata = prd, se.fit = TRUE)

prd$lci <- err$fit - 1.96 * err$se.fit
prd$fit <- err$fit
prd$uci <- err$fit + 1.96 * err$se.fit



# moderate_glm_model <- ggplot(goncalves_plot_data, aes(x = N_Infection, y=Severe_Moderate/Modelled_Supp_Cases))+
  ylab("\nRisk of Severe / Moderate Disease\n")+
  scale_y_continuous(label=scales::percent, limits=c(0, 0.15))+
  scale_x_continuous(limits=range(0, nrow(goncalves_plot_data)-1), breaks=seq(0, nrow(goncalves_plot_data)-1), expand = expansion(add = c(0.2,0.2)))+
  geom_point(colour="darkblue")+
  ggtitle("Severe / Moderate Disease")+
  theme_minimal()+
  geom_ribbon(data=prd, aes(x=N_Infection, ymin = exp(lci), ymax = exp(uci)),
              alpha = 0.2, inherit.aes = FALSE)+
  geom_function(fun = moderate_exp_fun, colour="blue")+
  theme(plot.title = element_text(size=15, hjust = 0.5),
        axis.text = element_text(size=12),
        axis.title = element_text(size=13.5))



pi <- summary(steady_decay_estimate)@coef[1]
pi_se <- summary(steady_decay_estimate)@coef[3]
alpha <- summary(steady_decay_estimate)@coef[2]
alpha_se <- summary(steady_decay_estimate)@coef[4]

prd2 <-  data.frame(N_Infection = seq(from = 1, to = 14, length.out = 100))
prd2$fit <- severe_constant_fun(prd2$N_Infection)

prd2$uci <- (pi+1.96*pi_se)-((alpha+1.96*alpha_se)*prd2$N_Infection)
prd2$lci <- (pi-1.96*pi_se)-((alpha-1.96*alpha_se)*prd2$N_Infection)
# prd2$uci <- (pi+(1.96*pi_se))-(alpha+(1.96*alpha_se))*prd2$N_Infection
# prd2$lci <- (pi-(1.96*pi_se))-(alpha-(1.96*alpha_se))*prd2$N_Infection


severe_constant_model <- ggplot(goncalves_plot_data, aes(x = N_Infection, y=Severe/Modelled_Main_Cases))+
  ylab("\nRisk of Severe Disease\n")+
  scale_y_continuous(label=scales::percent, limits=c(0, 0.075))+
  scale_x_continuous(limits=range(0, nrow(goncalves_plot_data)-1), breaks=seq(0, nrow(goncalves_plot_data)-1), expand = expansion(add = c(0.2,0.2)))+
  geom_point(colour="darkred")+
  ggtitle("Severe Disease")+
  theme_minimal()+
  geom_ribbon(data=prd2, aes(x=N_Infection, ymin = lci, ymax = uci),
              alpha = 0.2, inherit.aes = FALSE)+
  geom_function(fun = severe_constant_fun, color="red")+
  theme(plot.title = element_text(size=15, hjust = 0.5),
        axis.text = element_text(size=12),
        axis.title = element_text(size=13.5))



prd3 <- data.frame(N_Infection = seq(from = 1, to = 14, length.out = 100))
err3 <- predict(severe_glm, newdata = prd3, se.fit = TRUE)

prd3$lci <- err3$fit - 1.96 * err3$se.fit
prd3$fit <- err3$fit
prd3$uci <- err3$fit + 1.96 * err3$se.fit


severe_exponential_model <- ggplot(goncalves_plot_data, aes(x = N_Infection, y=Severe/Modelled_Main_Cases))+
  ylab("\nRisk of Severe Disease\n")+
  scale_y_continuous(label=scales::percent, limits=c(0, 0.075))+
  scale_x_continuous(limits=range(0, nrow(goncalves_plot_data)-1), breaks=seq(0, nrow(goncalves_plot_data)-1), expand = expansion(add = c(0.2,0.2)))+
  geom_point(colour="darkred")+
  ggtitle("Severe Disease")+
  theme_minimal()+
  geom_ribbon(data=prd3, aes(x=N_Infection, ymin = exp(lci), ymax = exp(uci)),
              alpha = 0.2, inherit.aes = FALSE)+
  geom_function(fun = severe_exp_fun, colour="red")+
  theme(plot.title = element_text(size=15, hjust = 0.5),
        axis.text = element_text(size=12),
        axis.title = element_text(size=13.5)
        #plot.margin = unit(c(2,2,2,2), "lines")
  )


plot(moderate_glm_model)
plot(severe_constant_model)
plot(severe_exponential_model)



```



